Verification of Stencil Computations through Spatial Specifications

### 2016-07-04 paper.tex
fix to effective dims
correct effective dimensions def
Make consistent with new syn & sem. Better syntax highlighting.

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}
\usepackage{amsthm}
\usepackage{tikz}
\usepackage{multirow}

@@

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation computes an array value, where the value at each index $i$ of
this array is calculated from a \emph{neighbourhood} of values around $i$ in
some input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following iteration computes the one-dimensional discrete Laplace
transform (an approximation to a derivative) in Fortran:
%
\begin{ExmVerbatim}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   end do
   a = b
end do
\end{ExmVerbatim}
%
Line 3 is the core of the stencil computation, calculating
the value at \texttt{b(i)} from a neighbourhood of elements about
\texttt{a(i)}. Line 4 swaps
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration. In Fortran, parentheses \texttt{( )} are used
for array subscripts rather than the more familiar bracket syntax \texttt{[ ]}.

@@

In this work, we introduce a simple specification language for the
spatial properties of stencil computations. The
specifications abstract over the fine grained detail of stencil access
patterns. In practice, most stencil computations have a regular shape
that can be described simply and abstractly, with a small set of
coarse-grained descriptions. In the case of our first Laplace example,
our inference procedure provides the specification:
%
\begin{SpecVerbatim}
!=  stencil centered(depth=1, dim=1) :: a
\end{SpecVerbatim}
%
This explains that \texttt{a} is read from with a
symmetrical stencil pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
In the case of the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Figure~\ref{ref:navier-stokes-fragment}(b). The
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Figure~\ref{ref:navier-stokes-fragment}(c) describes these
pictorially.

@@

\begin{figure}[t]
\begin{ExmVerbatim}[firstnumber=20]
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+
    gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))-
    (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-
    gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j)))
    /(4.0*delx)

duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+
   gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))-
   (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))-
   gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-u(i,j)))
   /(4.0*dely)

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+
          (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
\end{ExmVerbatim}
(a) Excerpt from Fortran code for a Navier-Stoke fluid simulator,
showing highly-detailed stencil computations. \\

@@

\begin{SpecVerbatim}[xleftmargin=0cm]
!= stencil centered(depth=1, dim=1)*reflexive(dim=2)
         + centered(depth=1, dim=2)*reflexive(dim=1) :: u

!= stencil forward(depth=1, dim=1)
         * backward(depth=1, dim=2) :: v
\end{SpecVerbatim}
(b) Inferred stencil specification from CamFort

@@

Regions have as terminals the \emph{region constants}, denoted by
\term{reflexive}, \term{forward}, \term{backward}, or \term{centered}. Each
region constant except \term{reflexive} is given a depth parameter $n$ (natural
number greater than 0). They all receive a dimension identifier $d$ (also a
natural number greater than 0). Region constants specify that an array is read
from by a collection of indices which in the $d^{th}$ dimension are
neighbourhood indices ranging from $i + 0$ up to $i \pm n$ inclusively,
\eg{}, the following is valid stencil, reading from \term{a} and writing to
\term{b}:
%%
\begin{ExmVerbatim}
!= stencil forward(depth=2, dim=1) :: a
b(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
\end{ExmVerbatim}
%%
were \texttt{i} is an induction variable.  The stencil specification
is associated to the array variable \texttt{a}. Note that the second
dimension is fixed at a constant here.

A \term{backward} stencil
is similar but has negative neighbourhood indices, \eg{},
%
\begin{ExmVerbatim}
!= stencil backward(depth=2, dim=1) :: a
b(i) = a(i) + a(i-1) + a(i-2)
\end{ExmVerbatim}
%
A \texttt{centered} stencil is a combination of \texttt{forward}
and \texttt{backward} stencils to the same depth, \eg{},
%%
\begin{ExmVerbatim}
!= stencil centered(depth=1, dim=1) :: a
b(i) = ( a(i) + a(i-1) + a(i+1) ) / 3.0
\end{ExmVerbatim}

@@

The sum of two regions $r \texttt{+} r'$ specifies that an array is
read using the neighbourhood indices described by either $r$ or
$r'$. It can be thought of as a disjunction of specifications. For
example, the following gives the specification for the five-point
stencil which is the sum of two \texttt{centered} regions in each
dimension:
%%
\begin{ExmVerbatim}
!= stencil centered(depth=1, dim=1)*reflexive(dim=2)
         + centered(depth=1, dim=2)*reflexive(dim=1) :: a

b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j)
                   + a(i,j-1) + a(i,j+1)
\end{ExmVerbatim}
%%
The product of two regions $r \texttt{*} r'$ specifies that the array
is read with neighbourhood indices drawn from the bounding box created
by the regions two regions $r$ and $r'$. For example, the following
code (defining a nine-point stencil) has a specification as the
product of two \texttt{centered} regions in each dimension:
%%
\begin{ExmVerbatim}
!= stencil centered(depth=1, dim=1)
         * centered(depth=1, dim=2) :: a

x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
b(i, j) = (x + y + z) / 9.0
\end{ExmVerbatim}
%%
Such a stencil computation is a common image convolution (for example
for edge detection). 
Note that there are a number of intermediate assignments over which
the whole specification ranges.

As another example, the following code and specification
implements the \emph{Roberts cross}
edge-detection convolution~\cite{davis1975survey}:
%%
\begin{ExmVerbatim}
do j=0, jmax-1
    do i=0, imax-1
\sten{      != region r1 = forward(depth=1, dim=1)}
\sten{      != region r2 = forward(depth=1, dim=2)}
\sten{      != stencil readOnce, r1*r2 :: a}
      x = a(i,j) - a(i+1,j+1)
      y = a(i+1,j) - a(i,j+1)
    b(i,j) = sqrt(x*x+y*y)
   end do
end do
\end{ExmVerbatim}

@@

Region specifications can be assigned to region variables via
a region declaration, which can be used later to form a spatial
specification. For example, the specification from the previous
 example can be restated as:
%%
\begin{ExmVerbatim}
!= region r1 = centered(depth=1, dim=1)
!= region r2 = centered(depth=1, dim=2)
!= stencil r1*r2 :: a
\end{ExmVerbatim}
%%
\paragraph{Spatial modifiers}
%%
There are three kinds of modifier that can be given to spatial
specifications: \texttt{readOnce}, \texttt{reflexive},
\texttt{irreflexive}.

The \texttt{readOnce} modifier enforces that no index appears
more than once. For example, in all of the previous examples the
\texttt{readOnce} modifier could be added, \eg{}
%
\begin{ExmVerbatim}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{ExmVerbatim}
%
However, this would be an invalid specification if any of the
array subscripts was repeated. This modifier provides a way to
 rule out any accidental repetition of array subscripts.
The notion corresponds to that of \emph{linearity}, though we
opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

The \texttt{irreflexive} modifier is parameterised by a
comma-separated list of dimensions, indicating that for those
dimensions there is no neighbour index with $0$ offset, \eg{},
%
\begin{ExmVerbatim}
!= region r = centered(depth=1, dim=1)
!= stencil irreflexive(dim=1), r :: a
b(i) = a(i+1) + a(i-1)
\end{ExmVerbatim}
%
or with just a \texttt{forward} region an example would be:
%
\begin{ExmVerbatim}
!= region r = forward(depth=1, dim=1)
!= stencil irreflexive(dim=1), r :: a
b(i) = a(i+1)
\end{ExmVerbatim}

The \term{reflexive} modifier is dual to \term{irreflexive},
indicating dimensions in which there is a neighbour index with $0$
offset. The region constants \term{forward}, \term{backward},
\term{centered} implicitly include the \term{reflexive} modifier,
but it is useful when there is a dimension in which there is reflexive
indexing but not a region constant, \eg{},
%%
\begin{ExmVerbatim}
!= region r = centered(depth=1, dim=1)
!= stencil reflexive(dims=2), r :: a
b(i,j) = a(i-1,j) + a(i+1,j) + a(i, j)
\end{ExmVerbatim}
%%
Overlapping \term{irreflexive} and \term{reflexive} modifiers
cannot occur together-- this is an invalid specification.

@@

In some cases, it is useful to give a lower and/or upper bound for a
stencil. This can be done using either the \term{atMost} or
\term{atLeast} modifiers. This is particularly useful in situations
where there is a non-contiguous stencil, which does not fit the rest
of the specification syntax.  For example:
%%
\begin{ExmVerbatim}
!= stencil atLeast, reflexive(dims=1)      :: a
!= stencil atMost, forward(depth=2, dim=1) :: a
b(i) = a(i) + a(i+2)
\end{ExmVerbatim}
%%
%\subsubsection{Temporal specifications}

@@

The names ``forward'', ``backward'' and ``centered''
are inspired by standard terminology in numerical analysis
for the shape of discretisation schemes. For example,
the standard ``explicit method'' for approximation
PDEs is the \emph{Forward Time, Centered Space} (FTCS)
scheme~\cite{dawson1991finite}. For
the one-dimensional heat equations, an FTCS discretisation
provides approximation code with the following stencil~\cite{recktenwald2004finite}:
%%
\begin{ExmVerbatim}
do i=2:n-1
  u(i) = r*v(i-1) + r2*v(i) + r*v(i+1)
end do
\end{ExmVerbatim}
%%
where \texttt{r} and \texttt{r2} are some constants.
A valid specification for this stencil is then:
%%
\begin{ExmVerbatim}
!= stencil centered(depth=1, dim=1)
         + centered(depth=1, dim=2) :: v
\end{ExmVerbatim}
%%
Such a specification can be inferred, or can be inserted by a user
and checked against the code.

@@

\begin{figure}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
\framebox{$\textit{region} \equiv \textit{region}'$} \\[1em]
\begin{array}{rll}
(\textsc{F\; \texttt{+} \;F}) \;\; &
\stenFwdS{n \, \textsf{max} \, m}{d} \\
 \equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenFwdS{m}{d} \\[1em]
%
(\textsc{B\; \texttt{+} \;B}) \;\; &
\stenBwdS{n \, \textsf{max} \, m}{d} \\
 \equiv \; & \stenBwdS{n}{d} \; \texttt{+} \; \stenBwdS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;C}) \;\; &
\stenCenS{n \, \textsf{max} \, m}{d} \\
\equiv \; & \stenCenS{n}{d} \; \texttt{+} \; \stenCenS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;F}) \;\; & \stenCenS{n}{d} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
                      \stenFwdS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;B}) \;\; &
\stenCenS{n}{d} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
                      \stenBwdS{m}{d} \\[1em]
(\textsc{B\; \texttt{+} \;F}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenBwdS{n}{d} \\[1em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.5em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.5em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.5em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.5em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T) \\[0.5em]
\end{array} \\ \\
%
\framebox{${\textit{specDec} :: \texttt{v}} \equiv
{\textit{specDec'} :: \texttt{v}}$} \\[1em]
\begin{array}{rl}
% MUTUAL
%(\textsc{mutual}) \; &
%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
%  :: \texttt{w}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
%  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; &
%\texttt{stencil} \; \texttt{dependency} (\bar{v})
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
%  :: \texttt{v}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
(\textsc{spatial}) \; &
(\textit{region} \equiv \textit{region'}) \\
\Rightarrow \; & (\texttt{stencil} \; [\textit{mod}] \; [\textit{approxMod},] \;
\textit{region} \\
& \!\!\equiv \; \texttt{stencil} \; [\textit{mod}] \;
            [\textit{approxMod},] \;
\textit{region}') \\
\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications}
\label{fig:equations}
\end{figure}

@@

\begin{definition}The product of two models $\otimes$ is defined:
%%
\begin{alignat*}{2}
& dims(M_m,N_n) =
  \bigg\{ x \; \bigg| \;
    \parbox{4.1cm}{
      $x \in \{1, \ldots, m\} \wedge M_x \neq \infty \; \vee \;$ \\
      $x \in \{1, \ldots, n\} \wedge N_x \neq \infty$
                  } \bigg\} \\
& M_n \otimes N_n =
  \Bigg\{x \; \Bigg| \;
    \parbox{4.8cm}{$i \in \{1, \ldots, 2^n\},$
                   $(u, v) \in M_n \times N_n,$ \\
                   $x = (u \bowtie v)_i,$ \\
                   $\forall j \in dims(M_n,N_n) (x_j \neq \infty) $
                  } \Bigg\}
\end{alignat*}
The $\bowtie$ operation, which we call \emph{pairwise permutation}
(described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
$n$-vectors). Using this, the $\otimes$ operation constructs the set of every
pairwise permutation for every pair $(u, v)$ of $n$-vectors in the Cartesian
product of $M \times N$.

@@

If the inference process cannot determine an exact stencil for a statement it will instead produce an upper (\term{atMost}) and lower (\term{atLeast}) bound. The numbers of exact and bounded stencils are shown in columns three and four of Figure \ref{fig:tickAssign}. We can see that our stencil language is capable of precisely specifying the overwhelming majority of stencils. Most of the bounded stencils apply to some form of data copying. For example (taken from E3MG):
\begin{ExmVerbatim}
!- stencil atLeast, readOnce, irreflexive(dims=2), (reflexive(dim=1)) :: sfda
!- stencil atMost, readOnce, (forward(depth=8, dim=2))*(reflexive(dim=1)) :: sfdt
ZZ1(I,J) = SFDT(I,J+8)
\end{ExmVerbatim}

>>>>

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}
\usepackage{amsthm}
\usepackage{tikz}
\usepackage{multirow}
\usepackage{minted}

\fvset{
  linenos=true,
  fontsize=\footnotesize,
  breaklines=true,
  breakafter=+*,
  xleftmargin=\parindent
}

@@

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation computes an array value, where the value at each index $i$ of
this array is calculated from a \emph{neighbourhood} of values around $i$ in
some input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following iteration computes the one-dimensional discrete Laplace
transform (an approximation to a derivative) in Fortran:
%
\begin{minted}{fortran}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   end do
   a = b
end do
\end{minted}
%
Line 3 is the core of the stencil computation, calculating
the value at \texttt{b(i)} from a neighbourhood of elements about
\texttt{a(i)}. Line 4 swaps
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration. In Fortran, parentheses \texttt{( )} are used
for array subscripts rather than the more familiar bracket syntax \texttt{[ ]}.

@@

In this work, we introduce a simple specification language for the
spatial properties of stencil computations. The
specifications abstract over the fine grained detail of stencil access
patterns. In practice, most stencil computations have a regular shape
that can be described simply and abstractly, with a small set of
coarse-grained descriptions. In the case of our first Laplace example,
our inference procedure provides the specification:
%
\begin{minted}{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
This explains that \texttt{a} is read from with a
symmetrical stencil pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
In the case of the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Figure~\ref{ref:navier-stokes-fragment}(b). The
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Figure~\ref{ref:navier-stokes-fragment}(c) describes these
pictorially.

@@

\begin{figure}[t]
  \begin{minted}[firstnumber=20]{fortran}
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+ &
  gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))- &
  gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) &
  /(4.0*delx)

duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+ &
  gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))- &
  (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))- &
  gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-u(i,j))) &
  /(4.0*dely)

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ &
  (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
\end{minted}
(a) Excerpt from Fortran code for a Navier-Stoke fluid simulator,
showing highly-detailed stencil computations. \\

@@

\begin{minted}{fortran}
!= stencil centered(depth=1,dim=1)*reflexive(dim=2) + centered(depth=1,dim=2)*reflexive(dim=1) :: u

!= stencil forward(depth=1,dim=1) * backward(depth=1,dim=2) :: v
\end{minted}
(b) Inferred stencil specification from CamFort

@@

Regions have as terminals the \emph{region constants}, denoted by
\term{reflexive}, \term{forward}, \term{backward}, or \term{centered}. Each
region constant except \term{reflexive} is given a depth parameter $n$ (natural
number greater than 0). They all receive a dimension identifier $d$ (also a
natural number greater than 0). Region constants specify that an array is read
from by a collection of indices which in the $d^{th}$ dimension are
neighbourhood indices ranging from $i + 0$ up to $i \pm n$ inclusively,
\eg{}, the following is valid stencil, reading from \term{a} and writing to
\term{b}:
%%
\begin{minted}{fortran}
!= stencil forward(depth=2, dim=1) :: a
b(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
\end{minted}
%%
were \texttt{i} is an induction variable.  The stencil specification
is associated to the array variable \texttt{a}. Note that the second
dimension is fixed at a constant here.

A \term{backward} stencil
is similar but has negative neighbourhood indices, \eg{},
%
\begin{minted}{fortran}
!= stencil backward(depth=2, dim=1) :: a
b(i) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
A \texttt{centered} stencil is a combination of \texttt{forward}
and \texttt{backward} stencils to the same depth, \eg{},
%%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1) :: a
b(i) = ( a(i) + a(i-1) + a(i+1) ) / 3.0
\end{minted}

@@

The sum of two regions $r \texttt{+} r'$ specifies that an array is
read using the neighbourhood indices described by either $r$ or
$r'$. It can be thought of as a disjunction of specifications. For
example, the following gives the specification for the five-point
stencil which is the sum of two compound \texttt{reflexive} and
\texttt{centered} regions in each dimension:
%%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1) :: a

b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j)
                   + a(i,j-1) + a(i,j+1)
\end{minted}
%%
Here the left hand side of \texttt{+} says when the second dimension
(induction variable $j$) is fixed at the origin, the first dimension
(induction variable $i$) accesses immediate vicinity of the origin.
The right hand side of \texttt{+} is similar but the dimensions are reversed.
This reflects the symmetry under rotation in the fivepoint stencil.

The product of two regions $r \texttt{*} r'$ specifies that the array
is read with neighbourhood indices drawn from the bounding box created
by the regions two regions $r$ and $r'$. For example, the following
code (defining a nine-point stencil) has a specification as the
product of two \texttt{centered} regions in each dimension:
%%
\begin{minted}[breakindent=2.9em]{fortran}
x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)

!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
b(i, j) = (x + y + z) / 9.0
\end{minted}
%%
Such a stencil computation is a common image convolution (for example
for edge detection).
Note that there are a number of intermediate assignments over which
the whole specification ranges.

The region constructor \texttt{*} can also be interpreted as conjunction of
two specifications since they need to hold simultaneously for their
corresponding dimensions.

As another example, the following code and specification
implements the \emph{Roberts cross}
edge-detection convolution~\cite{davis1975survey}:
%%
\begin{minted}[breakindent=5.5em]{fortran}
do j=0, jmax-1
    do i=0, imax-1
      x = a(i,j) - a(i+1,j+1)
      y = a(i+1,j) - a(i,j+1)
      != stencil forward(depth=1,dim=1) * forward(depth=1,dim=2) :: a
      b(i,j) = sqrt(x*x+y*y)
   end do
end do
\end{minted}

@@

Region specifications can be assigned to region variables via
a region declaration, which can be used later to form a spatial
specification. For example, the specification from the previous
 example can be restated as:
%%
\begin{minted}{fortran}
!= region r1 = centered(depth=1, dim=1)
!= region r2 = centered(depth=1, dim=2)
!= region robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is especially useful for common stencils such as Robert's cross
as the region can be defined once and used by region name many times.
%%
\paragraph{Spatial modifiers}
%%
There are two modifiers one used at specification the other at region
level to provide better descriptions for stencil computations.

At the specification level, the \texttt{readOnce} modifier enforces that no 
index appears more than once. For example, in all of the previous examples the
\texttt{readOnce} modifier could be added, \eg{}
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
However, this would be an invalid specification if any of the
array subscripts was repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion corresponds to that of \emph{linearity} in type systems,
though we opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

At the region level, \texttt{irreflexive} modifier qualifies base
regions of \texttt{forward}, \texttt{backward}, and \texttt{centered}.
It indicates that in the dimension of the region there is no neighbour index
with $0$ offset, \eg{},
%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1, irreflexive) :: a
b(i) = a(i+1) + a(i-1)
\end{minted}
%
or with just a \texttt{forward} region an example would be:
%
\begin{minted}{fortran}
!= stencil forward(depth=1, dim=1, irreflexive) :: a
b(i) = a(i+1)
\end{minted}

Naturally, \term{irreflexive} qualifier cannot be used on \term{reflexive}
regions.

@@

In some cases, it is useful to give a lower and/or upper bound for a
stencil. This can be done using either the \term{atMost} or
\term{atLeast} modifiers. This is particularly useful in situations
where there is a non-contiguous stencil, which does not fit the rest
of the specification syntax. For example:
%%
\begin{minted}{fortran}
!= stencil atLeast, reflexive(dim=1)      :: a
!= stencil atMost, forward(depth=2, dim=1) :: a
b(i) = a(i) + a(i+2)
\end{minted}
%%
%\subsubsection{Temporal specifications}

@@

The names ``forward'', ``backward'' and ``centered''
are inspired by standard terminology in numerical analysis
for the shape of discretisation schemes. For example,
the standard ``explicit method'' for approximation
PDEs is the \emph{Forward Time, Centered Space} (FTCS)
scheme~\cite{dawson1991finite}. For
the one-dimensional heat equations, an FTCS discretisation
provides approximation code with the following stencil~\cite{recktenwald2004finite}:
%%
\begin{minted}{fortran}
do i=2, n-1
  u(i) = r*v(i-1) + r2*v(i) + r*v(i+1)
end do
\end{minted}
%%
where \texttt{r} and \texttt{r2} are some constants.
A valid specification for this stencil is then:
%%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1) :: v
\end{minted}
%%
Such a specification can be inferred, or can be inserted by a user
and checked against the code.

@@

\begin{figure}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
\framebox{$\textit{region} \equiv \textit{region}'$} \\[1em]
\begin{array}{rll}
(\textsc{F\; \texttt{+} \;F}) \;\; &
\stenFwdS{n \, \textsf{max} \, m}{d} \\
 \equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenFwdS{m}{d} \\[1em]
%
(\textsc{B\; \texttt{+} \;B}) \;\; &
\stenBwdS{n \, \textsf{max} \, m}{d} \\
 \equiv \; & \stenBwdS{n}{d} \; \texttt{+} \; \stenBwdS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;C}) \;\; &
\stenCenS{n \, \textsf{max} \, m}{d} \\
\equiv \; & \stenCenS{n}{d} \; \texttt{+} \; \stenCenS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;F}) \;\; & \stenCenS{n}{d} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
                      \stenFwdS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;B}) \;\; &
\stenCenS{n}{d} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
                      \stenBwdS{m}{d} \\[1em]
(\textsc{B\; \texttt{+} \;F}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenBwdS{n}{d} \\[1em]
%
(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.5em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.5em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.5em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.5em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.5em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T) \\[0.5em]
\end{array} \\ \\
%
\framebox{${\textit{specDec} :: \texttt{v}} \equiv
{\textit{specDec'} :: \texttt{v}}$} \\[1em]
\begin{array}{rl}
% MUTUAL
%(\textsc{mutual}) \; &
%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
%  :: \texttt{w}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
%  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; &
%\texttt{stencil} \; \texttt{dependency} (\bar{v})
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
%  :: \texttt{v}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
(\textsc{spatial}) \; &
(\textit{region} \equiv \textit{region'}) \\
\Rightarrow \; & (\texttt{stencil} \; [\textit{mod}] \; [\textit{approxMod},] \;
\textit{region} \\
& \!\!\equiv \; \texttt{stencil} \; [\textit{mod}] \;
            [\textit{approxMod},] \;
\textit{region}') \\
\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications}
\label{fig:equations}
\end{figure}

@@

\begin{definition}The product of two models of dimensionality
$n$ is given by the tensor $\otimes_n$ defined:
%%
\newcommand{\effdims}[2]{\mathit{dims}(#1)_{#2}}
\begin{alignat*}{2}
& \effdims{M}{n} =
\bigcup_{u \in M} \{x \mid x \in \{1,...,n\}, u_x \neq \infty\} \\
& M \otimes_n N =
  \Bigg\{x \; \Bigg| \;
    \parbox{5.6cm}{$i \in \{1, \ldots, 2^n\},$
                   $(u, v) \in M \times N,$ \\
                   $x = (u \bowtie v)_i,$ \\
                   $j \in (\effdims{M}{n} \cup \effdims{N}{n}),$ \\
                   $x_j \neq \infty$
                  } \Bigg\}
\end{alignat*}
The $\bowtie$ operation, which we call \emph{pairwise permutation}
(described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
$n$-vectors). Using this, the $\otimes$ operation constructs the set of every
pairwise permutation for every pair $(u, v)$ of $n$-vectors in the Cartesian
product of $M \times N$.

@@

If the inference process cannot determine an exact stencil for a statement it will instead produce an upper (\term{atMost}) and lower (\term{atLeast}) bound. The numbers of exact and bounded stencils are shown in columns three and four of Figure \ref{fig:tickAssign}. We can see that our stencil language is capable of precisely specifying the overwhelming majority of stencils. Most of the bounded stencils apply to some form of data copying. For example (taken from E3MG):
\begin{minted}{fortran}
!- stencil atLeast, readOnce, irreflexive(dims=2), (reflexive(dim=1)) :: sfda
!- stencil atMost, readOnce, (forward(depth=8, dim=2))*(reflexive(dim=1)) :: sfdt
ZZ1(I,J) = SFDT(I,J+8)
\end{minted}
